Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-09-12 main.tex
Edits down to perturbations section.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles themselves are not self-stable
and can be challenging for the rider to balance. Low speed crashes may be
reduced if the bicycle was self-stable at these speeds by relieving the rider
of some of the necessary control activity. Bicycles can be mechanically
modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
and since the 1980s it has been known that applying a motor actuated steering
torque proportional to the roll rate can stabilize a single-track vehicle at
lower speeds than without the control. If automatic control of steering can
stabilize a bicycle, it may reduce the control required from the rider to
successfully manage balancing tasks. We have developed such a balance assisting
bicycle, Figure~\ref{fig:balance-assist-bicycle}, and hypothesize that it helps
the rider in situations in which they are likely to fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handledbar, which can easily cause a rider to
fall, especially when the rider is not so attentive or has one or more hands
removed from the handlebar.

In this paper, we test whether a linearly gain scheduled steer motor controlled
bicycle, that is stable in a large low speed range, is beneficial in helping to
prevent the rider from falling. We test this by applying varying magnitude
mechanical perturbations to the handlebars while the rider is cycling on a
treadmill. We assess the difference in the rider's probability of falling with
the balance assist steer motor system on and off.

\subsection{Background}
%
In the early years of developments in automatic control, \citep{Whipple1899}
not only derived the correct equations of motion of the bicycle but realized
and showed that roll motion feedback can stabilize a bicycle. Much later,
attempts at automatic roll stabilization of single track vehicles began in
earnest after predictive motorcycle models were developed and refined in
throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's work
and seems to be the first to attempt to robotically stabilize a small motorbike
with a controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} successfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
proposed how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities. Ruijs and Pacejka's work was not
particularly concerned with low speed stability and their vehicle was fully
automatic, no human rider was involved. Many more automatically balanced single
track vehicles have been demonstrated over the last 40 years, but none have
demonstrated that increasing low speed stability can assist in balancing and
possibly reduce single actor falls. Most of these robotic bicycle and
motorcycle designers did not intend for a human rider to also control the
stabilized vehicle. Nevertheless, an automatically stabilized bicycle can be
controlled by a human rider if the motor controlled steer torque and the rider
applied steer torque act on the steer in parallel.  The effect of this
automatic control gives the ability to change the dynamics, up to some limits,
of the human-controlled plant. In our prior study~\citep{Alizadehsaravi2023},
we demonstrate reduced motion during distractions due to the balance assist but
contrarily researchers with a similar vehicle recently showed rider
dissatisfaction with the stabilization~\cite{Hanakam2023}.

The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stability. The bicycle model is ideal for showing the effect of a roll
motion feedback driven steer motor on the dynamics. The linear version of this
model can be described by the fourth order state space
equations~\citep{Meijaard2007}:
%
\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
functions of the equilibrium forward speed \(v\) and  are otherwise populated
with expressions that are functions of the geometric and inertial parameters of
the nonholonomic multibody system made up of four rigid bodies: two wheels,
front frame, and rear frame.

@@

The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
such that the eigenvalues of \(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad
k_{\dot{\phi}} \quad 0 \quad 0\right] \right)\) have negative real parts for
\(v_{min} < v < v_\textrm{capsize}\) where \(v_{min}\) is the lowest stable
speed given \(k_{\dot{\phi}}\) and \(v_\textrm{capsize}\) is the speed at which
the uncontrolled bicycle's capsize mode goes unstable. With gain scheduling
with respect to \(v\), the speed range where the bicycle is stable can be
maximized within any physical actuator magnitude and bandwidth limits.
\citet{Schwab2008} elaborates on some of the possibilities in scheduling the
gains for such a controller and shows that a linear scheduling with respect to
speed can give satisfactory stability.

\section{Methods}
%
\subsection{Bicycle}
%
We modified an electric bicycle (Grenoble Arroyo\todo{Double check this is the
right model}, Royal Dutch Gazelle, Dieren, The Netherlands) with a custom motor
in the steering assembly capable of applying up to 7~\si{\newton\meter} of
torque between the head tube and steer tube, see
Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller converts the
commanded torque to applied torque. We measure the speed of the rear wheel with
an encoder (Magura XXX) and measure the body fixed roll rate of the bicycle
with a MEMs rate gyroscope (\todo{check bike for this}). The balance assist
control algorithm is implemented on a microprocessor (Teensy, PJRC, USA) and
data from all sensors is logged with a CAN bus (CanEdge2, CSS Electronics,
Denmark) at at least 100 Hz. \todo{Sample rate is not the same for each sensor:
wheel speed, imu = 1ms, steer torque command is updated every 10ms, can log is
written every 10ms}

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll rate feedback
controller to stabilize the bicycle. In the speed range tested, the commanded
steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the control
law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use \(g=8\) and \(g=10\) as gain values during the experiments. Scaling the
proportional feedback gain linearly with respect to speed stabilizes the
normally unstable weave mode of the bicycle down to
2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle  as shown in
Figure~\ref{fig:eigenvalues}~\todo{Add the experimental weave mode
identificatio to this figure and consider lowering the gain on the model to
make the model match the experimental values.}.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled (lower row) root locus of
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the bicycle without the inertia effect of a rigid rider (left) and with
    a rigid rider (right). Vertical blue lines indicate the two speeds we
    perform experiments at. Grey shaded region is the linear stable speed
    range. Geometric and inertial parameters for these plots were estimated
    with the methods presented in \cite{Moore2012} and software packages
    BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015}.}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in cooperation, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino Mega 2560, COMPANYXXX, Italy). We measure the force
applied by each motor at the handlebar via four inline 250~\si{\newton} load
cells. The left rear load cell was damaged during pilot experiments, and is
replaced by a load cell with a max load of 500~\si{\newton}. The commanded
force profiles are designed to apply an external pulsive torque to the front
assembly (handlebars, forks, wheel) at magnitudes varying from 16 to
160~\si{\newton\meter}. The four motors are arranged at the four corners of a
1~\si{\meter} wide treadmill (EC-90 Flat, Maxon Group, Switzerland) that can
reach speeds of 18~\si{\kilo\meter\per\hour}. The general design of this system
is described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
Our modifications relative to Van De Velde include simplifying the controller
with a inexpensive microcontroller and the use of a simpler non-active safety
harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[width=75mm]{figures/participant-in-set-up.png}}
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[width=65mm]{figures/setup-diagram.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations.}
  \label{fig:experimental-setup}
\end{figure}

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles themselves are not self-stable
and can be challenging for the rider to balance. Low speed crashes may be
reduced if the bicycle was self-stable at these speeds by relieving the rider
of some of the necessary control activity. Bicycles can be mechanically
modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
and since the 1980s it has been known that applying a motor actuated steering
torque proportional to the vehicle's roll rate can stabilize a single-track
vehicle at lower speeds. If automatic control of steering can stabilize a
bicycle, it may reduce the control required from the rider to successfully
manage balancing tasks much like the high speed self-stability already does. We
have developed a balance assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handledbar, which can relatively easily cause
a rider to fall.

In this paper, we test whether a steer motor controlled bicycle, that is stable
in a large low speed range, is beneficial in helping to prevent the rider from
falling. We test this by applying varying magnitude mechanical perturbations to
the handlebars while the rider is cycling on a treadmill. We assess the
difference in the rider's probability of falling with the balance assist system
on and off.

\subsection{Technical Background}
%
During the early years of developments in automatic control, \citep{Whipple1899}
not only derived the correct equations of motion of the bicycle but realized
and showed that roll motion feedback can stabilize a bicycle. Much later,
attempts at automatic roll stabilization of single-track vehicles began in
earnest after predictive motorcycle models were developed and refined in
throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's work
and seems to be the first to attempt to robotically stabilize a small motorbike
with a controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} successfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so solely with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
proposed how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback enables the simplest controller
that can stabilize a single-track vehicle above a minimum speed when one is not
concerned with wobble instabilities. Ruijs and Pacejka's work was not
particularly concerned with low speed stability and their vehicle was fully
automatic, i.e no human rider was involved.

Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single-actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to effectively change
the dynamics, up to some limits, of the human-controlled plant. In our prior
study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
distractions due to the balance assist system but contrarily researchers with a
similar vehicle recently showed rider dissatisfaction with the
stabilization~\cite{Hanakam2023}.

The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stability. The bicycle model is suitable for showing the effect of a roll
motion feedback driven steer motor on the dynamics. The linear version of this
model~\citep{Meijaard2007} can be described by the fourth order state space
equations:
%
\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
\(T_\delta\). The state \(\mathbf{A}\) matrix is a function of the equilibrium
forward speed \(v\). It and the input \(\mathbf{B}\) matrix are otherwise
populated with expressions that are functions of the geometric and inertial
parameters of the nonholonomic multibody system made up of four rigid bodies:
two wheels, front frame, and rear frame.

@@

The state matrix \(\mathbf{A}\) being a function of the equilibrium speed \(v\)
means \(k_{\dot{\phi}}\) can be selected such that the eigenvalues of
\(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad k_{\dot{\phi}} \quad 0 \quad
0\right] \right)\) have negative real parts for \(v_{min} < v <
v_\textrm{capsize}\) where \(v_{min}\) is the lowest stable speed given
\(k_{\dot{\phi}}\) and \(v_\textrm{capsize}\) is the speed at which the
uncontrolled bicycle's capsize mode goes unstable. With gain scheduling with
respect to \(v\), the speed range where the bicycle is stable can be maximized
within any physical actuator magnitude and bandwidth limits.
\citet{Schwab2008} elaborates on some of the possibilities in scheduling the
gains for such a controller and shows that a linear scheduling with respect to
speed can give satisfactory stability for a low speed range. We use this simple
feedback principle as the basis for our balance assist controller.

\section{Methods}
%
\subsection{Bicycle}
%
We modified an electric bicycle (Grenoble Arroyo, Royal Dutch Gazelle, Dieren,
The Netherlands) with a custom motor in the steering assembly capable of
applying up to 7~\si{\newton\meter} of torque between the head tube and steer
tube, see Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller
converts the commanded torque to applied torque. We measure the speed of the
rear wheel with an encoder (Magura XXX) and measure the body fixed roll rate of
the bicycle with a MEMs rate gyroscope (BN0086, Sparkfun, Niwot, USA). The
balance assist control algorithm is implemented on a microprocessor (Teensy,
PJRC, USA) and data from all sensors is logged with a CAN bus (CanEdge2, CSS
Electronics, Denmark) at at least 100 Hz.

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll rate feedback
controller to stabilize the bicycle. In the speed range tested, the commanded
steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the control
law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use \(g=8\) and \(g=10\) as gain values during the experiments. Scaling the
proportional feedback gain linearly with respect to speed stabilizes the
normally unstable weave mode of the bicycle down to
2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle  as shown in
Figure~\ref{fig:eigenvalues}~\todo{Could add the experimental weave mode
identification to this figure and consider lowering the gain on the model to
make the model match the experimental values.}.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled (lower row) root locus of
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the bicycle without the inertial effect of a rigid rider (left) and
    with a rigid rider (right). Vertical blue lines indicate the two speeds we
    perform experiments at. Grey shaded region is the linear stable speed
    range. Geometric and inertial parameters for these plots were estimated
    with the methods presented in \cite{Moore2012} and software packages
    BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015}.}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in cooperation, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino Mega 2560, Arduino, Italy). We measure the force
applied by each motor at the handlebar via four inline 250~\si{\newton} load
cells. The left rear load cell was damaged during pilot experiments, and is
replaced by a load cell with a max load of 500~\si{\newton}. The commanded
force profiles are designed to apply an external pulsive torque to the front
assembly (handlebars, forks, wheel) at magnitudes varying from 16 to
160~\si{\newton\meter}. The four motors are arranged at the four corners of a
1~\si{\meter} wide treadmill (EC-90 Flat, Maxon Group, Switzerland) that can
reach speeds of 18~\si{\kilo\meter\per\hour}. The general design of this system
is described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
Our modifications relative to Van De Velde include simplifying the controller
with a inexpensive microcontroller and the use of a simpler non-active safety
harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[width=75mm]{figures/participant-in-set-up.png}}
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[width=65mm]{figures/setup-diagram.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations.}
  \label{fig:experimental-setup}
\end{figure}
